An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-09-23 nips_2016.tex

Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta'-\theta,u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated M-H tests in~\cite{cutting_mh_2014,icml2014c1_bardenet14} require at
least a constant fraction $\geq c$ in the amount of data consumed per test
compared to full-dataset tests, so their speed-up is at most $1/c$.
{\color{blue} Daniel: doesn't our method also have this issue? In general, it
seems like any test that requires minibatch information to show deviation from
the full batch version will suffer from this problem. Why not just force us to
stop if we use too much data?}

@@

Assume we begin with the current sample $\theta$ and a candidate sample
$\theta'$, and that $V \sim \mathcal{U}(0,1)$ is a uniform random variable. We
accept $\theta'$ if $g(\Delta(\theta,\theta')) > V$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is
well-defined and unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}\label{eq:equivalent_test}
    \Delta(\theta,\theta') > X = g^{-1}(V)
\end{equation}
where $X$ is a random variable with the logistic distribution (its CDF is the
logistic function). To see this notice that $\frac{dV}{dX} = g'$, that $g'$ is
the density corresponding to a logistic CDF, and finally that $\frac{dV}{dX}$ is
the density of $X$. The density of $X$ is symmetric, so we can equivalently test
whether
\begin{equation}\label{eq:the_exact_test}
    \Delta(\theta,\theta') + X > 0
\end{equation}
for a logistic random variable $X$. {\color{blue} Daniel: I think the text
before Equation~\ref{eq:equivalent_test} should be similar to what we had in
our NIPS submission because I find it much clearer.}

@@

Our approach is to instead decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
    X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm} \sim \mathcal{N}(0, \sigma^2)$ and $X_{\rm corr}$
is a zero-mean ``correction'' variable with density $C_{\sigma}(X)$.  The two
variables are added (i.e., their distributions convolve) to form $X_{\rm log}$.
This decomposition requires an appropriate $X_{\rm corr}$ distribution. We show
in Section~\ref{sec:correction} that it is possible to use deconvolution to
derive a highly accurate representation of the density $C_\sigma(X)$.
Figure~\ref{fig:deconvolution} shows three examples of $X_{\rm norm}$ and
$X_{\rm corr}$ distributions that sum to $X_{\rm log}$.

@@

If $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ is exactly $\mathcal{N}(0,
\sigma^2(\Delta^*))$, the above test is exact as well, and as we show in
Section~\ref{sec:analysis}, if there is a maximum error $\epsilon$ between the
CDF of $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ and the CDF of $\mathcal{N}(0,
\sigma^2(\Delta^*))$, then the acceptance test has an error of at most
$\epsilon$ relative to the full batch version. {\color{blue} Daniel: this
subsection is improved from earlier. My main concern now is that $X_{\rm norm}$
is described as being both approximately Gaussian $\bar{\mathcal{N}}$ and
Gaussian $\mathcal{N}$ here. We may want to clear that up; in our NIPS
submission we actually used a new variable $\varepsilon$ for that and assigned
it to be the $\mathcal{N}$ distribution.}

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, error bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial sample $\theta_1$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ from $p(\theta)$\;}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution $q(\theta'\mid\theta_t)$\;
    Draw a minibatch of $m$ points $x_i$, compute $\Delta^*(\theta_t,\theta')$ and sample variance $\sigma^2(\Delta^*)$\;
    Estimate moments $E|\Lambda_i-\Lambda|$ and $E|\Lambda_i-\Lambda|^3$ from the sample, and error $\epsilon$ from equation (\ref{eq:clt-bounds3})\;
    \While {$\sigma^2(\Delta^*) \ge 1$ {\bf or} $\epsilon > \delta$}{
    	Draw $m$ more samples to augment the minibatch, update $\Delta^*$, $\sigma^2(\Delta^*)$ and $\epsilon$ estimates\;
    }
    Draw $X_{\rm norm} \sim \mathcal{N}(0,1-\sigma^2(\Delta^*))$ and
    $X_{\rm corr}$ from the correction distribution $C_1(X)$\;

@@

}
\caption{A description of M-H sampling with our acceptance test. {\color{blue}
Daniel: I think we may want to be careful about the use of $C_1(X)$ because that
implies that we have a fixed $X_{\rm corr}$ density, but that $C_1(X)$ will
change depending on the iteration and variance of $\Delta^*$, right? Hopefully
not by much, but it still changes sometimes?}}
\label{alg:our_algorithm}
\end{algorithm}

@@

In the most similar prior works,~\cite{cutting_mh_2014} uses asymptotic
arguments based on the Central Limit Theorem to argue that its approximate
acceptance test error tends to zero as batch size increases. But no quantitative
bounds are given. In~\cite{icml2014c1_bardenet14}, explicit bounds are given,
but they depend on bounding
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
Such bounds can be derived efficiently for models such as logistic regression,
but it is unclear how to derive them for a complex model such as a neural
network. In general, since we obtain a new $\theta'$ value each iteration, one
would need to iterate through all the $p(x_i\mid \theta')$ terms, defeating the
purpose of minibatch MCMC.

@@

{\color{blue} Daniel: the notation here was (and still is, to a lesser extent)
the source of considerable confusion for me. Do we want $\sigma(X)$ or
$\sigma(\hat{X})$? I think in Lemma~\ref{lem:quant_clt} we want the former, not
the latter. OH ALSO, this section uses $n$ as the full data size, which means
before this section, we should use $n$, not $N$, for the full data size, right?
The Bardenet paper uses $n$ to represent the full data size.}

@@

>>>>

Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta'-\theta,u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated M-H tests in~\cite{cutting_mh_2014,icml2014c1_bardenet14} require at
least a constant fraction $\geq c$ in the amount of data consumed per test
compared to full-dataset tests, so their speed-up is at most $1/c$.

{\color{blue} Daniel: quick question: I think we should explain why our method
\emph{doesn't} run into this issue. I think it is because we rely on an
approximation to $\Delta + X_{\rm log} > 0$ and that test does not require any
``difference between an approximation and an exact version'' because it is
\emph{already} an exact version, assuming the approximation is enough.

Also, in terms of this example, if we are in the special case with using a lot
of data, why not force early stopping?}

@@

Assume we begin with the current sample $\theta$ and a candidate sample
$\theta'$, and that $V \sim \mathcal{U}(0,1)$ is a uniform random variable. We
accept $\theta'$ if $g(\Delta(\theta,\theta')) > V$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is
well-defined and unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}\label{eq:equivalent_test}
    \Delta(\theta,\theta') > X = g^{-1}(V)
\end{equation}
where $X$ is a random variable with the logistic distribution (its CDF is the
logistic function). To see this notice that $\frac{dV}{dX} = g'$, that $g'$ is
the density corresponding to a logistic CDF, and finally that $\frac{dV}{dX}$ is
the density of $X$. The density of $X$ is symmetric, so we can equivalently test
whether
\begin{equation}\label{eq:the_exact_test}
    \Delta(\theta,\theta') + X > 0
\end{equation}
for a logistic random variable $X$. {\color{blue} Daniel: I think the text
before Equation~\ref{eq:the_exact_test} should be similar to what we had in our
NIPS submission because I find it much clearer.}

@@

Our approach is to instead decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
    X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm} \sim \mathcal{N}(0, \sigma^2)$ and that $X_{\rm
corr}$ is a zero-mean ``correction'' variable with density $C_{\sigma}(X)$.  The
two variables are added (i.e., their distributions convolve) to form $X_{\rm
log}$.  This decomposition requires an appropriate $X_{\rm corr}$ distribution.
We show in Section~\ref{sec:correction} that it is possible to use deconvolution
to derive a highly accurate representation of $C_\sigma(X)$.
Figure~\ref{fig:deconvolution} shows three examples of $X_{\rm norm}$ and
$X_{\rm corr}$ densities that sum to $X_{\rm log}$.

@@

If $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ is exactly $\mathcal{N}(0,
\sigma^2(\Delta^*))$, the above test is exact as well, and as we show in
Section~\ref{sec:analysis}, if there is a maximum error $\epsilon$ between the
CDF of $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ and the CDF of $\mathcal{N}(0,
\sigma^2(\Delta^*))$, then the acceptance test has an error of at most
$\epsilon$ relative to the full batch version.

{\color{blue} Daniel: this subsection is improved from earlier. My main concern
now is that $X_{\rm norm}$ is described as being both approximately Gaussian
$\bar{\mathcal{N}}$ and Gaussian $\mathcal{N}$ here. We may want to clear that
up; in our NIPS submission we actually used a new variable $\varepsilon$ for
that and assigned it to be the $\mathcal{N}$ distribution. My current
suggestion is to have $\Delta^*$ be the $\bar{\mathcal{N}}$ approx. normal
distribution (but with mean $\Delta$, of course), and keep $X_{\rm norm}$ as an
exact Gaussian.}

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, error bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial sample $\theta_1$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ from $p(\theta)$\;}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution $q(\theta'\mid\theta_t)$\;
    Draw a minibatch of $m$ points $x_i$, compute $\Delta^*(\theta_t,\theta')$ and sample variance $\sigma^2(\Delta^*)$\;
    Estimate moments $E|\Lambda_i-\Lambda|$ and $E|\Lambda_i-\Lambda|^3$ from the sample, and error $\epsilon$ from Equation~\ref{eq:clt-bounds3}\;
    \While {$\sigma^2(\Delta^*) \ge 1$ {\bf or} $\epsilon > \delta$}{
    	Draw $m$ more samples to augment the minibatch, update $\Delta^*$, $\sigma^2(\Delta^*)$ and $\epsilon$ estimates\;
    }
    Draw $X_{\rm norm} \sim \mathcal{N}(0,1-\sigma^2(\Delta^*))$ and
    $X_{\rm corr}$ from the correction distribution $C_1(X)$\;

@@

}
\caption{A description of M-H sampling with our acceptance test. {\color{blue}
Daniel: I think we may want to be careful about the use of $C_1(X)$ because that
implies that we have a fixed $X_{\rm corr}$ density, but that $C_1(X)$ will
change depending on the iteration and variance of $\Delta^*$, right? Hopefully
not by much, but it still changes sometimes? ALSO bring up Xinlei's old points
about $X_{\rm norm}$ confusion.}}
\label{alg:our_algorithm}
\end{algorithm}

@@

In the most similar prior works,~\cite{cutting_mh_2014} uses asymptotic
arguments based on the Central Limit Theorem to argue that its approximate
acceptance test error tends to zero as batch size increases. But no quantitative
bounds are given. In~\cite{icml2014c1_bardenet14}, explicit bounds are given,
but they depend on bounding
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
Such bounds can be derived efficiently for models such as logistic regression,
but it is unclear how to derive them for a complex model such as a neural
network. In general, since a new $\theta'$ value is obtained each iteration, one
would need to iterate through all the $p(x_i\mid \theta')$ terms, defeating the
purpose of minibatch MCMC.

@@

{\color{blue} Daniel: the notation here was (and still is, to a lesser extent)
the source of considerable confusion for me. Do we want $\sigma(X)$ or
$\sigma(\hat{X})$? I think in Lemma~\ref{lem:quant_clt} we want the former, not
the latter. OH ALSO, this section uses $n$ as the full data size, which means
before this section, we should use $n$, not $N$, for the full data size, right?
The Bardenet paper uses $n$ to represent the full data size.

PS: I'd recommend not putting sample mean and sample variance in an equation if
the $\sigma(\hat{X})$ gets clarified, since those computations are common
knowledge.}

@@

{\color{blue} Daniel: Hold on, the following Lemma seems to add $X_{\rm norm}$
to both random variables we are comparing (and \emph{then} $X_{\rm corr}$). But
didn't the previous section already justify that $\Delta^* \approx \Delta +
X_{\rm norm}$? That is, I don't think we need to add another $X_{\rm norm}$; I
think we just want to add $X_{\rm corr}$. Otherwise, this new test does not
match what we wrote earlier.}
